---
title: "lab1"
author: "204814891_204169320"
date: "14 4 2020"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	message = FALSE,
	warning = FALSE
)
library('maps')
library('ggplot2')
library('mapproj')
library('nycflights13')
library('readr')
library('dplyr')
```


We loaded a  bunch of libararys that will help us later.
this includes then nycflights13 library that has the following data frames that we will use:
- flights 
- airlines 
- airports 
- planes 
- weather


Q1 

After we did some exloration (by ourselfes) of the data, let's answer Q 1:

1.1)
Delayed flights fig:
   This graphic is telling a clear story, it is telling us the percent of flights that are delayed by more then 15 minutes, from Denver Colorado to other destinatinos in the US. 
   
Weekly Cycles fig:

Q2
seif a:
- first thing we need is the number of flights, and number of delade flights each day

```{r}
flights$date <- format(flights$time_hour, '%D')
clean_flights <- na.omit(flights)
clean_flights$delayed <- ifelse(clean_flights$dep_delay > 15, 1, 0)
flights_per_day <- as.data.frame(table(factor(clean_flights$date)))
colnames(flights_per_day) <- c("date", "flights")
delayes_per_day <- clean_flights %>% group_by(date) %>% summarise(Tot.delayed = sum(delayed))
colnames(delayes_per_day) <- c("date", "delayes")
data4graph <- merge(flights_per_day, delayes_per_day)
data4graph$date <- as.Date(data4graph$date, "%m/%d/%Y")
```
After setting up all the data in the way we want it, all we need to do is plot!

```{r}
ggplot(data4graph) + geom_line(aes(x=date,y= flights, color = 'All flights (schedualed for separtue)')) + geom_bar(stat = "identity", aes(x = date, y = delayes, color = 'Late flights (departure delayed > 15 minutes)')) + theme(legend.position = "bottom") + scale_color_manual(name = "Colors", values = c('All flights (schedualed for separtue)' = "blue", 'Late flights (departure delayed > 15 minutes)' = "red")) + labs(title = "Weekly Cycles: Flights from NYC airports in 2013 \n  Few flights: Sundays & Saturdays \n Most Flights: Mondays, thursdays & fridays", y = "Flights per day") + theme(plot.title = element_text(hjust = 0.5))
```
Q2
seif b
```{r}
# > 15 min. delay
delays <- subset(flights, dep_delay > 15)
del_dest_count <- table(delays$dest)
dest_overall <- table(flights$dest)
```

We can see that we have 3 airports withot delays at all:
```{r}
dest_overall[!(names(table(flights$dest)) %in% names(table(delays$dest)))]
```
So we'll ignore them and then assign them manually 0%
```{r}
no_del <- rownames(dest_overall[!(names(table(flights$dest)) %in% names(table(delays$dest)))])
dest_overall <- dest_overall[names(table(flights$dest)) %in% names(table(delays$dest))]
# compute percentages
perc_delays <- as.data.frame(del_dest_count / dest_overall, row.names = TRUE)

# append and assign
to_append <- data.frame(rep(0,3))
colnames(to_append) <- 'Freq'
rownames(to_append) <- no_del
perc_delays <- rbind(perc_delays,to_append)
```

```{r}
# add perc. bins according to the example
perc_delays$bin = cut(perc_delays$Freq, 
                  c(0,0.1,0.15,0.20,0.25,1), right = FALSE, 
                  labels=c("<= 10%",
                           "10% - 15%",
                           "15% - 20%",
                           "15% - 25%",
                           "> 25%"))

#define JFK as the source (all 3 are pretty close in output graph)
perc_delays[,c('lon_source','lat_source')] <- subset(airports, faa=='JFK', select = c('lon','lat'))

```
during getting target coordinates, we discovered 4 airports which don't have coordinated in the airports dataframe:
```{r}
rownames(perc_delays)[!(rownames(perc_delays) %in% airports$faa)]
```
After Googling it we found they all located in Puerto Riko, so we could omit them relying on the assumption that the relevant data is located only in the main land of the U.S.A (same as Denver's plot in Moodle).

```{r}
perc_delays <- perc_delays[rownames(perc_delays) %in% airports$faa,]
```
Now, it's possible to get the target coordinates
```{r}
perc_delays[,c('lon_target','lat_target')] <- airports[airports$faa %in% rownames(perc_delays),c('lon','lat')]
```

plot
```{r message=FALSE, warning=FALSE}
USA <- subset(map_data("world"), region=="USA")
#USA <- subset(USA, lat < 50) remove alaska
states <- read_csv("states.csv") # state name with location
```



```{r}
plot <- 
  ggplot() +
  # plot usa with state borders and names
  geom_polygon(data = USA, aes(x=long, y = lat, group = group), fill="white", alpha=0.3) +
  borders("state") +
  geom_text(aes(x=longitude,y=latitude, label=state), data=states, size = 2) +
# plot filghts according to delays
geom_segment(data = perc_delays, aes(x=lon_source,y=lat_source, xend=lon_target, yend=lat_target, colour=bin)) +
geom_point(data=perc_delays,aes(x=lon_target,y=lat_target, colour=bin),shape=21,size=1.5, show.legend = FALSE) +

scale_fill_manual(values=rep("white",5)) + 
scale_colour_manual(values=c("forestgreen", "purple", "blue","orange","red"))
```

```{r warning=FALSE}
plot +
  ggtitle("% of Flights Departures Delayed > 15 Min", subtitle = "airport=NY's airports    Year=2013") +

  theme(legend.title = element_blank(), legend.position = c(0.22,0.1), panel.background = element_blank(), axis.line = element_blank(), axis.title = element_blank(), axis.ticks = element_blank(), axis.text = element_blank(),legend.text = element_text(size=7), legend.key = element_rect(fill = "transparent", colour = "transparent"), legend.key.height = unit(0.7,"line"), plot.title = element_text(hjust = 0.5, size=10, face="bold"), plot.subtitle = element_text(hjust = 0.5, size=6, face="bold")) +

    xlim(-130,-70) + 
    ylim(25,50) +
    coord_map()
```

Q3
We want to examine how departures delays depend on the flight's day of week and the daytime (the latter could be examine dependenty and independenty with the former).
Firstly, we'll subset the relevant data.In addition, we'll create a daytime bins in order to make the plot more interpretable.
```{r}
data <- select(flights,year,month,day,hour,dep_delay,time_hour)
data$daytime <- cut(data$hour, c(0,4,11,18,23),labels=c("evening (18h-01h)","morning (05h-11h)","afternoon (11h-18h)","evening (18h-01h)"),right = TRUE)
```

the data doesn't include dat of week column, we can use the built-in weekdays functions, but we have to convert the date format beforhand.
```{r}
# function to convert date data to one string with dash sep
date_as_str <- function(year,month,day){
  if(any(is.na(c(year,month,day)))){
    return(NA)
  }
  year <- as.character(year)
  ifelse(month<10, paste('0',as.character(month),sep=''),
         as.character(month))
  ifelse(day<10, paste('0',as.character(day),sep=''),
         as.character(day))
    return(paste(year,month,day,sep='-'))
}

```

Now, we can create the new desired column
```{r message=FALSE}
Sys.setlocale(category = "LC_ALL", locale = "English_United States.1252")
data$day_of_week <- weekdays(as.Date(date_as_str(data$year,data$month,data$day)))
data$day_num_of_week <- sapply(data$day_of_week,switch,"Sunday"=1,"Monday"=2,"Tuesday"=3, "Wednesday"=4,"Thursday"=5,"Friday"=6,"Saturday"=7)
```

We'll create a df summarises the average delays.
```{r}
week_sum <- data %>%
  group_by(day_num_of_week) %>%
  summarise(average_delay = mean(dep_delay, na.rm = 1))
```

Now, we'll create a summary for average delay per hour for each day of the week.
```{r}
hour_week_sum <- data %>%
  group_by(day_num_of_week, daytime) %>%
  summarise(average_delay = mean(dep_delay, na.rm = 1))
```


```{r}
ggplot(data=hour_week_sum) +
  ggtitle("Departures Delays per Day of Week", subtitle = "Dots Represents the Avg Delay per Daytime in a Day") +
  labs(x="Day", y="Avg Delay (minutes)") +
  
  # per daytime
  geom_point(aes(x=day_num_of_week,y=average_delay,colour=daytime), show.legend = TRUE) +
  
  # daily
  geom_point(data = week_sum,aes(x=day_num_of_week,y=average_delay, fill="daily avg"), size=2.5) +
  labs(fill="") +
  geom_line(data = week_sum,aes(x=day_num_of_week,y=average_delay)) +
  geom_text(data=week_sum, aes(x=day_num_of_week,y=average_delay+2, label=round(average_delay,digits = 2)),fontface='bold',size=4) +
    
  scale_x_continuous(breaks = c(1:7), labels = c("Sun","Mon","Tues","Wed","Thu","Fri","Sat"))
```

As for the days, it looks like the Saturday is more likely to has a minor delay while Monday and Thursday are more likely to have some more major delays. The interesting thing is that could be seen nicely from the plot, is that no matter what day it is, you should prefer choose a morning flight and not an evening or an afternoon flight in terms of departure delays.